%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% negram.tex
%
% diana, 8/1
%
% $Id: negram.tex,v 1.9 2005/02/10 17:08:59 ian Exp $
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapt[chap:negram]{Named-Entity State Machine Patterns}
\markboth{Named-Entity State Machine Patterns}{Named-Entity State Machine
Patterns}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%% qqqqqqqqqqqqqqqqqqqqqqqqq %%%%
\begin{quote}
There are, it seems to me, two basic reasons why minds aren't computers...
The first... is that human beings are organisms. Because
of this we have all sorts of needs - for food, shelter, clothing, sex
etc - and capacities - for locomotion, manipulation, articulate
speech etc, and so on - to which there are no real analogies in computers.
These needs and capacities underlie and interact with our
mental activities. This is important, not simply because we can't
understand how humans behave except in the light of these needs
and capacities, but because any historical explanation of how human
mental life developed can only do so by looking at how this
process interacted with the evolution of these needs and capacities
in successive species of hominids. 

\ldots

The second reason... is that... brains don't work like computers. 

{\it Minds, Machines and Evolution}, Alex Callinicos, 1997 (ISJ 74, p.103).
\end{quote}
%%%% qqqqqqqqqqqqqqqqqqqqqqqqq %%%%


This \chapthing\ describes the individual grammars used in GATE for Named
Entity Recognition, and how they are combined together. It relates to
the default NE grammar for ANNIE, but should also provide guidelines
for those adapting or creating new grammars. For documentation about
specific grammars other than this core set, use this document in
combination with the comments in the relevant grammar
files. \chapthing\ \ref{chap:jape} also provides information
about designing new grammar rules and tips for ensuring maximum
processing speed.


\sect{Main.jape}

This file contains a list of the grammars to be used, in the
correct processing order. The ordering of the grammars is crucial,
because they are processed in series, and later grammars may depend on
annotations produced by earlier grammars.

The default grammar consists of the following phases:
\begin{itemize}
\item first.jape
\item firstname.jape
\item name.jape
\item name\_post.jape
\item date\_pre.jape
\item date.jape
\item reldate.jape
\item number.jape
\item address.jape
\item url.jape
\item identifier.jape
\item jobtitle.jape
\item final.jape
\item unknown.jape
\item name\_context.jape
\item org\_context.jape
\item loc\_context.jape
\item clean.jape
\end{itemize}


\sect{first.jape}

This grammar must always be processed first. It can contain any
general macros needed for the whole grammar set. This should consist
of a macro defining how space and control characters are to be
processed (and may consequently be different for each grammar set,
depending on the text type). Because this is defined first of all, it
is not necessary to restate this in later grammars. This has a big
advantage -- it means that default grammars can be used for specialised
grammar sets, without having to be adapted to deal with e.g. different
treatment of spaces and control characters. In this way, only the
first.jape file needs to be changed for each grammar set, rather than
every individual grammar.

The first.jape grammar also has a dummy rule in. This is never
intended to fire -- it is simply added because every grammar set must
contain rules, but there are no specific rules we wish to add
here. Even if the rule were to match the pattern defined, it is
designed not to produce any output (due to the empty RHS).

\sect{firstname.jape}
This grammar contains rules to identify first names and titles via the
gazetteer lists. It adds a gender feature where appropriate from the
gazetteer list. This gender feature is used later in order to improve
co-reference between names and pronouns. The grammar creates
separate annotations of type FirstPerson and Title.

\sect{name.jape}

This grammar contains initial rules for organization, location and
person entities. These rules all create temporary annotations, some of
which will be discarded later, but the majority of which will be
converted into final annotations in later grammars. Rules beginning
with `Not' are negative rules -- this means that we detect something
and give it a special annotation (or no annotation at all) in order to
prevent it being recognised as a name. This is because we have no
negative operator (we have `=' but not `!=').

\subsect{Person}

We first define macros for initials, first names, surnames, and
endings. We then use these to recognise combinations of first names
from the previous phase, and surnames from their POS tags
or case information. Persons get marked with the
annotation `TempPerson'. We also percolate feature information about
the gender from the previous annotations if known.

\subsect{Location}

The rules for Location are fairly straightforward, but we define them
in this grammar so that any ambiguity can be resolved at the top
level. Locations are often combined with other entity types, such as
Organisations. This is dealt with by annotating the two entity types
separately, and them combining them in a later phase. Locations are
recognised mainly by gazetteer lookup, using not only lists of known
places, but also key words such as mountain, lake, river, city
etc. Locations are annotated as TempLocation in this phase.

\subsect{Organization}

Organizations tend to be defined either by straight lookup from
the gazetteer lists, or, for the majority, by a combination of POS or
case information and key words such as `company', `bank',
`Services' `Ltd.' etc. Many organizations are also identified by
contextual information in the later phase org\_context.jape. In this
phase, organizations are annotated as TempOrganization.

\subsect{Ambiguities}

Some ambiguities are resolved immediately in this grammar, while
others are left until later phases. For example, a Christian name
followed by a possible Location is resolved by default to a person rather than
a Location (e.g. `Ken London'). On the other hand, a Christian name
followed by a possible organisation ending is resolved to an
Organisation (e.g. `Alexandra Pottery'), though this is a slightly
less sure rule.

\subsect{Contextual information}
Although most of the rules involving contextual information are
invoked in a much later phase, there are a few which are invoked here,
such as `X joined Y' where X is annotated as a Person and Y as an
Organization. This is so that both annotations types can be handled at once.

\sect{name\_post.jape}
This grammar runs after the name grammar to fix some erroneous
annotations that may have been created. Of course, a more elegant
solution would be not to create the problem in the first instance, but
this is a workaround. For example, if the surname of a Person contains
certain stop words, e.g. `Mary And' then only the first name should be
recognised as a Person. However, it might be that the firstname is
also an Organization (and has been tagged with TempOrganization
already), e.g. `U.N.' If this is the case, then the annotation is left
untouched, because this is correct.

\sect{date\_pre.jape}
This grammar precedes the date phase, because it includes extra context
to prevent dates being recognised erroneously in the middle of longer
expressions. It mainly treats the case where an expression is already
tagged as a Person, but could also be tagged as a date (e.g. 16th
Jan).

\sect{date.jape}
This grammar contains the base rules for recognising times and
dates. Given the complexity of potential patterns representing such
expressions, there are a large number of rules and macros.

Although times and dates can be mutually ambiguous, we try to
distinguish between them as early as possible. Dates, times and years
are generally tagged separately (as TempDate, TempTime and TempYear
respectively) and then recombined to form a final Date annotation in a
later phase. This is because dates, times and years can be combined
together in many different ways, and also because there can be much
ambiguity between the three. For example, 1312 could be a time or a
year, while 9-10 could be a span of time or date, or a fixed time
or date.

\sect{reldate.jape}
This grammar handles relative rather than absolute date and time
sequences, such as `yesterday morning', `2 hours ago', `the first
9 months of the financial year'etc. It uses mainly explicit key words
such as `ago' and items from the gazetteer lists.


\sect{number.jape}

This grammar covers rules concerning money and percentages.
The rules are fairly straightforward, using keywords from the
gazetteer lists, and there is little ambiguity here, except for
example where `Pound' can be money or weight, or where there is no
explicit currency denominator.

 
\sect{address.jape}

Rules for Address cover ip addresses, phone and fax
numbers, and postal addresses. In general, these are not highly ambiguous, and can be
covered with simple pattern matching, although phone numbers can
require use of contextual information. Currently only UK formats are
really handled, though handling of foreign zipcodes and
phone number formats is envisaged in future. The annotations produced
are of type Email, Phone etc. and are then replaced in a later phase
with final Address annotations with `phone' etc. as features.

\sect{url.jape}
Rules for email addresses and Urls are in a separate grammar from the
other address types, for the simple reason that SpaceTokens need to be
identified for these rules to operate, whereas this is not necessary
for the other Address types. For speed of processing, we place them in
separate grammars so that SpaceTokens can be eliminated from the Input
when they are not required.

\sect{identifier.jape}

This grammar identifies `Identifiers' which basically means any
combination of numbers and letters acting as an ID, reference number
etc. not recognised as any other entity type. 

\sect{jobtitle.jape}

This grammar simply identifies Jobtitles from the gazetteer lists, and
adds a JobTitle annotation, which is used in later phases to aid
recognition of other entity types such as Person and Organization. It
may then be discarded in the Clean phase if not required as a final
annotation type.


\sect{final.jape}

This grammar uses the temporary annotations previously assigned in
the earlier phases, and converts them into final annotations. The
reason for this is that we need to be able to resolve ambiguities
between different entity types, so we need to have all the different
entity types handled in a single grammar somewhere. Ambiguities can be
resolved using prioritisation techniques. Also, we may need to combine
previously annotated elements, such as dates and times, into a single
entity.

The rules in this grammar use Java code on the RHS to
remove the existing temporary annotations, and replace them with new
annotations. This is because we want to retain the features associated
with the temporary annotations. For example, we might need to keep
track of whether a person is male or female, or whether a location is
a city or country. It also enables us to keep track of which rules
have been used, for debugging purposes.

For the sake of obfuscation, although this phase is called final, it
is not the final phase!

\sect{unknown.jape}

This short grammar finds proper nouns not previously recognised,
and gives them an Unknown annotation. This is then used by the
namematcher -- if an Unknown annotation can be matched with a
previously categorised entity, its annotation is changed to that of
the matched entity. Any remaining Unknown annotations are useful
for debugging purposes, and can also be used as input for additional
grammars or processing resources.


\sect{name\_context.jape}

This grammar looks for Unknown annotations occurring in certain
contexts which indicate they might belong to Person. This is a typical
example of a grammar that would benefit from learning or automatic
context generation, because useful contexts are (a) hard to find
manually and may require large volumes of training data, and (b) often
very domain--specific. In this core grammar, we confine the use of
contexts to fairly general uses, since this grammar should not be
domain--dependent.

\sect{org\_context.jape}

This grammar operates on a similar principle to name\_context.jape. It
is slightly oriented towards business texts, so does not quite fulfil
the generality criteria of the previous grammar. It does, however,
provide some insight into more detailed use of contexts.</p>

\sect{loc\_context.jape}

This grammar also operates in a similar manner to the preceding two,
using general context such as coordinated pairs of locations, and
hyponymic types of information.

\sect{clean.jape}

This grammar comes last of all, and simply aims to clean up (remove)
some of the temporary annotations that may not have been deleted along
the way.

%\include{grammars}







