%'nested' switch is used to make section headings format correctly whether
%they are in the appendix or the introduction.
%
% Within this file you should use \rcSect[label]{Heading} instead of
% \sect[sec:changes:label]{Heading}, and \rcSubsect{Heading} instead of
% \subsect{Heading}.  These commands are redefined appropriately depending on
% whether we are processing the intro subsection or the full changelog
% appendix.
%
\ifnested
  \def\rcSectNoLabel#1{\subsect{#1}}
  \def\rcSect[#1]#2{\subsect[subsec:changes:#1]{#2}}
  \def\rcSubsect#1{\subsubsect{#1}}
  \def\rcSubsubsect#1{\subsubsubsect{#1}}
\else
  \def\rcSectNoLabel#1{\sect{#1}}
  \def\rcSect[#1]#2{\sect[sec:changes:#1]{#2}}
  \def\rcSubsect#1{\subsect{#1}}
  \def\rcSubsubsect#1{\subsubsect{#1}}
\fi

\rcSect[9.0.1]{Version 9.0.1 (March 2021)}
GATE Developer 9.0.1 is a bugfix release -- the only change is to the way URL redirects
are handled when loading a document. Support for following redirects from http to https
was added in 9.0 which, while correct, broke the way URLs were used within GCP. This
release fixes that bug and adds some additional security checking to the redirect handling.

\rcSect[9.0]{Version 9.0 (February 2021)}

Whilst the majority of changes in GATE Developer 9.0 are small a number of them
change default behaviour (in the UI or API) hence the change in version number.
These changes include:

\begin{itemize}
\item We now recommend users install a 64 bit version of Java whenever possible. This
      seems to be especially important on Windows.
\item We now default to assuming documents are UTF-8 encoded unless you specify otherwise.
      In previous versions if no encoding was specified GATE would use the default platform
      encoding, but this seemed to cause more problems than it solved (especially for Windows
      users). If you want the old behaviour then ensure the encoding parameter is set to the
      empty string when creating a document.
\item GATE uses a library called XStream for saving and loading GATE XML documents and
      applications. This allows us to store features of any Java type, but that can be
      abused by maliciously crafted files. In general use this is unlikely to be a problem,
      but in situations where GATE may be used as part of a service with no way of vetting
      input files it could present a serious security threat. XStream now offers a security
      framework to restrict the types of objects that can be loaded/saved. This can work either
      by allowing only specific types or by preventing specific types from being used. As we
      often do not know in advance what features might be used we have opted to use a minimal
      blacklist as the default security setting. This blocks the Java classes known to be
      exploitable. This can be further configured via calls to \verb!Gate.setXStreamSecurity()!
      and we strongly encourage developers who depend on gate-core within larger applications
      to configure this based on their specific use cases.
\item Developers wishing to build GATE from source need to use Maven v3.6.0 or above.
\item Previous versions of GATE used Log4J for some of the logging. This was problematic
      when using gate-core as a dependency in larger projects and was awkward to configure
      properly. In this release we've switched to using SLF4J allowing the actual logging
      back-end to be configured independently. Plugins and code compiled against previous
      versions of GATE should work with the new release without change (we include the
      log4j-over-slf4j bridge as a dependency), although Log4J specific methods within
      gate-core have been deprecated and may be removed in a future release.
\end{itemize}

Many bugs have been fixed and documentation improved, in particular:

\begin{itemize}
\item the \verb!Twitter! plugin has been improved to make better use of the information
      provided by Twitter within a JSON Tweet object. The Hashtag tokenizer has been
      updated to provide a \verb!tokenized! feature to make grouping semantically similar
      hashtags easier. Lots of other minor improvements and efficiency changes have been
      made throughout the rest of the TwitIE pipelines.
\item the \verb!ANNIE! gazetteers have been updated to better support different ways of referring
      to countries and a blacklist option to prevent things being wrongly annotated.
\item A new addition to the JAPE syntax allows you to copy all features from a matched annotation
      to the new annotation being created
\item the \verb!Format_CSV! plugin now allows the document cell to be interpreted as being
      a URL pointing to the document to load rather than the contents of the document. See
      Section~\ref{sec:creole:csv} for more details.
\end{itemize}

\rcSect[8.6.1]{Version 8.6.1 (January 2020)}

GATE Developer 8.6.1 is a bugfix release -- the only change is to adjust for
the fact that the Central Maven repository has been switched from \verb!http!
to \verb!https!.

\rcSect[8.6]{Version 8.6 (June 2019)}

GATE Developer 8.6 is mainly a maintenance and stability release, but there are
some important new features, in particular around the processing of
Twitter data:

\begin{itemize}
\item The \verb!Format_Twitter! plugin can now correctly handle extended 280
  character tweets and the latest Twitter JSON format. See
  Section~\ref{sec:social:twitter:format} for full details.
\item The new \verb!Format_JSON! plugin provides import/export support for GATE
  JSON.  This is essentially the old style Twitter format, but it no longer
  needs to track changes to the Twitter JSON format so should be more suitable
  for long term storage of GATE documents as JSON files. See
  Section~\ref{sec:creole:gatejson} for more details.  This plugin makes use of
  a new mechanism whereby document format parsers can take parameters via the
  document MIME type, which may be useful to third party formats too.
\end{itemize}

Many bugs have been fixed and documentation improved, in particular:

\begin{itemize}
\item The plugin loading mechanism now properly respects the user's Maven
  \verb!settings.xml!:
  \begin{itemize}
  \item HTTP proxy and ``mirror'' repository settings now work properly,
  including authentication.  Also plugin resolution will now use the system
  proxy (if there is one) by default if there is no proxy specified in the
  Maven settings.
  \item The ``offline'' setting is respected, and will prevent GATE from trying
  to fetch plugins from remote repositories altogether -- for this to work, all
  the plugins you want to use must already be cached locally, or you can use
  ``Export for GATE Cloud'' to make a self-contained copy of an application
  including all its plugins.
  \end{itemize}
\item Upgraded many dependencies including Tika and Jackson to avoid known
  security bugs in the previous versions.
\item Documentation improvements for the Kea plugin, the Corpus QA and
  annotation diff tools, and the default GATE XML and inline XML formats
  (section~\ref{sec:developer:dump})
\item For plugin developers, the standard plugin testing framework generates a
  report detailing all the plugin-to-plugin dependencies, including those that
  are only expressed in the plugin's example saved applications
  (section~\ref{sec:api:bootstrap:dependencies}).
\end{itemize}

Some obsolete plugins have been removed (Websphinx web crawler, which depends
on an unmaintained library, and the RASP parser, whose external binary is no
longer available for modern operating systems), and there are many smaller bug
fixes and improvements.

Note: following changes to Oracle's JDK licensing scheme, we now recommend
running GATE using the freely-available OpenJDK.  The
\htlink{https://adoptopenjdk.net}{AdoptOpenJDK} project offers simple
installers for all major platforms, and major Linux distributions such as
Ubuntu and CentOS offer OpenJDK packages as standard.  See
section~\ref{sec:gettingstarted:install} for full installation instructions.

\rcSect[8.5.1]{Version 8.5.1 (June 2018)}

Version 8.5.1 is a minor release to fix a few critical bugs in 8.5:

\begin{itemize}
\item Fixed an exception that prevented the ANNIC search GUI from opening.
\item Fixed a problem with ``Export for GATE Cloud'' that meant some resources
  were not getting included in the output ZIP file.
\item Fixed the XML schema in the \verb!gate-spring! library.
\end{itemize}

\rcSect[8.5]{Version 8.5 (May 2018)}

GATE Developer and Embedded 8.5 introduces a number of significant internal
changes to the way plugins are managed, but with the exception of the plugin
manager most users will not see significant changes in the way they use GATE.

\begin{itemize}
\item The GATE plugins are no longer bundled with the GATE Developer
  distribution, instead each plugin is downloaded from a repository at runtime,
  the first time it is used.  This means the distribution is much smaller than
  previous versions.
\item Most plugins are now distributed as a single JAR file through the
  Java-standard ``Central Repository'', and resource files such as gazetteers
  and JAPE grammars are bundled inside the plugin JAR rather than being
  separate files on disk.  If you want to modify the resources of a plugin
  then GATE provides a tool to extract an editable copy of the files from
  a plugin onto your disk -- it is no longer possible to edit plugin grammars
  in place.
\item This makes dependencies between plugins much easier to manage -- a plugin
  can specify its dependencies declaratively by name and version number rather
  than by fragile relative paths between plugin directories.
\end{itemize}

GATE 8.5 remains backwards compatible with existing third-party plugins, though
we encourage you to convert your plugins to the new style where possible.

Further details on these changes can be found in sections
\ref{sec:developer:plugins} (the plugin manager in GATE Developer),
\ref{sec:api:plugins} (loading plugins via the GATE Embedded API),
\ref{sec:api:bootstrap} (creating a new plugin from scratch), and
\ref{sec:api:updateplugins} (converting an existing plugin to the new style).

If you have an existing saved application from GATE version 8.4.1 or earlier it
will be necessary to ``upgrade'' it to use the new core plugins.  An upgrade
tool is provided on the ``Tools'' menu of GATE Developer, and is described in
section Section~\ref{sec:developer:convertxgapp}.

\rcSubsect{For developers}

As part of this release, GATE development has moved from SourceForge to GitHub
-- bug reports, patches and feature requests should now use the GitHub issue
tracker as described in section~\ref{sec:development:bugs}.

% vim:ft=tex
